PAMELA: an annotation-based Java Modeling Framework

### 2021-02-02 architecture.tex
work on architecture and process (AB)

Figure \ref{fig:PamelaVision} summarizes the PAMELA architecture.  Basically, it is composed of one design time component (the PAMELA metamodel, plus a number of predefined annotations) presented in the next section and one run-time component (the PAMELA interpreter), described in the section \ref{sub:RunTime}.

\begin{figure}
    \centering
    \includegraphics[width=1.0 \columnwidth]{PamelaVisionV2.pdf}
    \caption{PAMELA approach for modelling}
    \label{fig:PamelaVision}
\end{figure}

@@

The PAMELA metamodel is depicted in Figure  \ref{fig:PamelaMetaModel}, and allows, though the use of annotations, the definition of of \emph{simple} UML-like models directly on the code.

A \texttt{PAMELAModel} is defined as a set of references to \texttt{ModelEntity}. A \texttt{ModelEntity} reflects a concept and is encoded in a java \texttt{interface}. PAMELA metamodel allows multiple inheritance: thus \texttt{ModelEntity} may define a set of parent entities. Note that reification of \texttt{ModelEntity} is performed in a java \texttt{interface} (and not a class), which only defines API whithout any implementation for methods. 
Note that \emph{ModelEntities} may declare \emph{ImplementationClasses}, which will be responsible for providing custom code implementing domain or application specific behaviour. A partially implemented \texttt{abstract} java \texttt{class} may be defined as partial base implementation (conforms to implemented interface).

A \texttt{ModelEntity} also defines some properties, encoded as \texttt{ModelProperty}. A \texttt{ModelProperty} is identified by a name, a cardinality (simple or multiple) and a type, which can be a reference to another \texttt{ModelEntity}, or a Java type (a primitive or an arbitrary complex Java type). Depending on its cardinality, a \texttt{ModelProperty} is bound to a set of methods reflecting use of property.
\begin{itemize}
    \item A \emph{read-only single property} will define read-access of its value using a \emph{getter} (a java method defined in java interface taking no argument and returning desired value).
    \item A \emph{read-write single property} will define a \emph{getter} and a \emph{setter} (a java method taking value to be set as unique argument)
    \item A \emph{read-write multiple property} will define a \emph{getter}, an \emph{adder} (a java method taking value to be added as unique argument), a \emph{remover} (a java method taking value to be removed as unique argument), and may define additional methods for extended features such as reindexing for example.
\end{itemize}

To summarize, in the following we list the most common java annotations used in the context of \texttt{ModelEntity} and \texttt{ModelProperty} definitions:

\begin{itemize}
    \item \texttt{@ModelEntity}: tag annotating \texttt{interface} as \emph{ModelEntity}. May declare an abstract entity.
    \item \texttt{@ImplementationClass}: tag annotating \emph{ModelEntity} \texttt{interface} and precising abstract java \texttt{class} to be used as base implementation.
    \item \texttt{@Implementation}: tag annotating a partial implementation (abstract inner \texttt{class} defined in implemented \texttt{interface}), and used in the context of multiple inheritance.
    \item \texttt{@Getter(String)}: tag annotating method as unique getter for implicit \emph{ModelProperty} whose identifier is the declared String value. May also declares cardinality, enventual inverse property, default value and some other features.
    \item \texttt{@Setter(String)}: tag annotating method as unique setter for implicit \emph{ModelProperty} whose identifier is the declared String value.
    \item \texttt{@Adder(String)}: tag annotating method as unique adder for implicit multiple cardinality \emph{ModelProperty} whose identifier is the declared String value.
    \item \texttt{@Remover(String)}: tag annotating method as unique remover for implicit multiple cardinality \emph{ModelProperty} whose identifier is the declared String value.
    \item \texttt{@Reindexer(String)}: tag annotating method as unique reindexer for implicit multiple cardinality \emph{ModelProperty} whose identifier is the declared String value.
    \item \texttt{@Initializer}: tag annotating a method used as a constructor for related \emph{ModelEntity}
   \item \texttt{@Deleter}: tag annotating a method used as explicit destructor for related \emph{ModelEntity}
    \item \texttt{@Finder(String,String)}: tag annotating method as a fetching request for a given \emph{ModelProperty} with a given attribute
    \item \texttt{@CloningStrategy}: allows to customize cloning strategy for a given \emph{ModelProperty}
    \item \texttt{@Embedded}: allows to declare a given \emph{ModelProperty} as embedded according to PAMELA semantics
    \item \texttt{@Imports} and \texttt{@Imports}: allows to declare some entities to be included in PAMELA model
    \item \texttt{@XMLElement} and \texttt{@XMLAttribute}: used to specify XML serialization for PAMELA instances

@@

 The aforementioned models are executed at runtime as a combination of two components: 1) plain java byte-code, as the result of the basic compilation of source code; and 2) an embedded PAMELA interpreter, executing semantics reflected by \emph{ModelEntity} and \emph{ModelProperty}  declarations (together with custom annotations where available).

The main idea for the approach is to override java dynamic binding. Invoking a method on an object which is part of a PAMELA model, causes the real implementation to be called when existing (more precisely dispatch code execution between all provided implementations), or the required interpretation according to the underlying model to be executed.

@@

PAMELA model at runtime is computed dynamically, working on the classpath of launched java application, and starting from a simple java interface (or a collection of java interfaces) which is/are PAMELA-annotated. From a mathematical point of view, internal representation of the underlying model is a graph whose vertex are PAMELA \emph{ModelEntities} (annotated java interface), and edges are either inheritance links or reference links (a property whose type is another \emph{ModelEntity}. \mytexttt{@Imports} and \mytexttt{@Import} annotations allows to include some other \emph{ModelEntities} in the model. On the contrary, an annotation attribute \mytexttt{@Getter(...ignoreType=true)} allows to ignore the link. In that context, PAMELA model computation is a graph closure computation, starting from a collection of vertices\footnote{Model closure computation on-the-fly provides an interesting approach to deal with model fragmentation.}. A PAMELA model at runtime is represented by a \mytexttt{ModelContext}.

@@

% Folloging section/itemize may be removed when needed 
%This composition offers many benefits: 
%\begin{itemize}
%    \item Strong coupling between model and code
%    \item Strong typing is kept, and required checks are performed by the java compiler
%    \item PAMELA framework provides interpretation of model@runtime
%    \item No need to generate POJO (plain old java objects), as their execution follow the standard semantics (less code, less bugs)
%    \item Custom implementation are provided if needed, using classical java extension points
%    \item It offers a way to intercept method calls and instrument the code
%    \item Assertions checking at runtime
%    \item Dynamic code weaving at runtime (aspect programming without compilation)
%\end{itemize}

@@

PAMELA supports a number of advanced programming features (some already mentioned) such as:  multiple inheritance and traits, containment management, cloning, fine-grained notifications management, object graph comparison and diff/merge, visiting patterns, clipboard management, validation, support for \emph{design by contract} by integrating the Java Modeling Language (JML), Metaprogramming, Aspect Oriented Programming.

%A strong interest of the approach is that the model is encoded in java, and must be compiled. It forces the java compiler to perform required checks %for a PAMELA model encoded in a strong typed program. Execution semantics of model is fully compatible with Java semantics. Many validation rules %are automatically performed through classical java compilation, independently of underlying PAMELA execution semantics.

@@

Listing \ref{lst:model} shows a very basic model with two entities: \emph{Book} and \emph{Library}. Entity \emph{Book} defines two read-write single properties \emph{title} and \emph{ISBN} with single cardinality and with \texttt{String} type. Entity \emph{Book} also defines a constructor with initial \emph{title} value. Entity \emph{Library} defines a read-write multiple properties \emph{books} referencing \emph{Book} instances. Note that this code is sufficient to execute the model, while no additional line of code is required (only java interfaces and API methods are declared here).

@@

The first line of code instantiates a \texttt{ModelContext} by introspecting and computing the closure of concepts graph obtained while starting from \texttt{Library} entity and following \texttt{parentEntities} and \texttt{properties} relationships. This call builds at runtime a \emph{PAMELAModel}, while dynamically following links reflected by compiled byte-code. A factory \texttt{ModelFactory} is then instantiated using that \texttt{ModelContext}, allowing to create \emph{Library} and \emph{Book} instances.

Custom code can be easily added to this model as we show in Listing \ref{lst:custom}. It shows how to integrate custom code to the fully interpreted \emph{Book} entity described above. The partial custom implementation is offered by a partial class (note the \texttt{abstract} keyword), declared in the annotation header of model entity. Custom implementations are defined using classical java implementation/overrides scheme. Here we define the implementation of the \texttt{read()} method, which has no annotation (and thus, can not be processed by the PAMELA framework), and also the implementation of a custom getter for \emph{title}, returning a default value when no value is defined for that property. Note that this implementation references a default interpretated implementation (call to \texttt{performSuperGetter(String)} method).

>>>>

PAMELA is composed of one design time component (the PAMELA metamodel, plus a number of predefined annotations) presented in the next section and one run-time component (the PAMELA interpreter), described in the section \ref{sub:RunTime}.

\subsection{Installation}
\label{sub:installation}

\todo{Maybe a word on how to install.}

@@

The PAMELA metamodel is depicted in Figure  \ref{fig:PamelaMetaModel}, and allows, though the use of annotations, the definition of \emph{simple} UML-like models directly on the code.

A \texttt{PAMELAModel} is defined as a set of references to \texttt{ModelEntity}. A \texttt{ModelEntity} reflects a concept and is encoded in a Java \texttt{interface}. PAMELA metamodel allows multiple inheritance: thus \texttt{ModelEntity} may define a set of parent entities. Note that reification of \texttt{ModelEntity} is performed in a Java \texttt{interface} (and not a class), which only defines API whithout any implementation for methods. 
Note that \emph{ModelEntities} may declare \emph{ImplementationClasses}, which will be responsible for providing custom code implementing domain or application specific behaviour. A partially implemented \texttt{abstract} Java \texttt{class} may be defined as partial base implementation (conforms to implemented interface).

A \texttt{ModelEntity} also defines some properties, encoded as \texttt{ModelProperty}. A \texttt{ModelProperty} is identified by a name, a cardinality (simple or multiple) and a type, which can be a reference to another \texttt{ModelEntity}, or a Java type (a primitive or an arbitrary complex Java type). Depending on its cardinality, a \texttt{ModelProperty} is bound to a set of methods reflecting use of property.
\begin{itemize}
    \item A \emph{read-only single property} will define read-access of its value using a \emph{getter} (a Java method defined in Java interface taking no argument and returning desired value).
    \item A \emph{read-write single property} will define a \emph{getter} and a \emph{setter} (a Java method taking value to be set as unique argument).
    \item A \emph{read-write multiple property} will define a \emph{getter}, an \emph{adder} (a Java method taking value to be added as unique argument), a \emph{remover} (a Java method taking value to be removed as unique argument), and may define additional methods for extended features such as reindexing for example.
\end{itemize}

To summarize, in the following we list the most common Java annotations used in the context of \texttt{ModelEntity} and \texttt{ModelProperty} definitions:

\begin{itemize}
    \item \texttt{@ModelEntity}: tag annotating \texttt{interface} as \emph{ModelEntity}. May declare an abstract entity.
    \item \texttt{@ImplementationClass}: tag annotating \emph{ModelEntity} \texttt{interface} and precising abstract Java \texttt{class} to be used as base implementation.
    \item \texttt{@Implementation}: tag annotating a partial implementation (abstract inner \texttt{class} defined in implemented \texttt{interface}), and used in the context of multiple inheritance.
    \item \texttt{@Getter(String)}: tag annotating method as unique getter for implicit \emph{ModelProperty} whose identifier is the declared String value. May also declares cardinality, eventual inverse property, default value and some other features.
    \item \texttt{@Setter(String)}: tag annotating method as unique setter for implicit \emph{ModelProperty} whose identifier is the declared String value.
    \item \texttt{@Adder(String)}: tag annotating method as unique adder for implicit multiple cardinality \emph{ModelProperty} whose identifier is the declared String value.
    \item \texttt{@Remover(String)}: tag annotating method as unique remover for implicit multiple cardinality \emph{ModelProperty} whose identifier is the declared String value.
    \item \texttt{@Reindexer(String)}: tag annotating method as unique reindexer for implicit multiple cardinality \emph{ModelProperty} whose identifier is the declared String value.
    \item \texttt{@Initializer}: tag annotating a method used as a constructor for related \emph{ModelEntity}.
   \item \texttt{@Deleter}: tag annotating a method used as explicit destructor for related \emph{ModelEntity}.
    \item \texttt{@Finder(String,String)}: tag annotating method as a fetching request for a given \emph{ModelProperty} with a given attribute.
    \item \texttt{@CloningStrategy}: allows to customize cloning strategy for a given \emph{ModelProperty}.
    \item \texttt{@Embedded}: allows to declare a given \emph{ModelProperty} as embedded according to PAMELA semantics.
    \item \texttt{@Imports} and \texttt{@Imports}: allows to declare some entities to be included in PAMELA model.
    \item \texttt{@XMLElement} and \texttt{@XMLAttribute}: used to specify XML serialization for PAMELA instances.

@@

 The aforementioned models are executed at runtime as a combination of two components: 1) plain Java byte-code, as the result of the basic compilation of source code; and 2) an embedded PAMELA interpreter, executing semantics reflected by \emph{ModelEntity} and \emph{ModelProperty}  declarations (together with custom annotations where available).

The main idea for the approach is to override Java dynamic binding. Invoking a method on an object which is part of a PAMELA model, causes the real implementation to be called when existing (more precisely dispatch code execution between all provided implementations), or the required interpretation according to the underlying model to be executed.

@@

PAMELA model at runtime is computed dynamically, working on the classpath of launched Java application, and starting from a simple Java interface (or a collection of Java interfaces) which is/are PAMELA-annotated. From a mathematical point of view, internal representation of the underlying model is a graph whose vertex are PAMELA \emph{ModelEntities} (annotated Java interface), and edges are either inheritance links or reference links (a property whose type is another \emph{ModelEntity}. \mytexttt{@Imports} and \mytexttt{@Import} annotations allows to include some other \emph{ModelEntities} in the model. On the contrary, an annotation attribute \mytexttt{@Getter(...ignoreType=true)} allows to ignore the link. In that context, PAMELA model computation is a graph closure computation, starting from a collection of vertices.
%\footnote{Model closure computation on-the-fly provides an interesting approach to deal with model fragmentation.} 
A PAMELA model at runtime is represented by a \mytexttt{ModelContext}.

@@

% Folloging section/itemize may be removed when needed 
%This composition offers many benefits: 
%\begin{itemize}
%    \item Strong coupling between model and code
%    \item Strong typing is kept, and required checks are performed by the Java compiler
%    \item PAMELA framework provides interpretation of model@runtime
%    \item No need to generate POJO (plain old Java objects), as their execution follow the standard semantics (less code, less bugs)
%    \item Custom implementation are provided if needed, using classical Java extension points
%    \item It offers a way to intercept method calls and instrument the code
%    \item Assertions checking at runtime
%    \item Dynamic code weaving at runtime (aspect programming without compilation)
%\end{itemize}

@@

\ab{Programmers may already reuse} a number of advanced programming features (some already mentioned) such as:  multiple inheritance and traits, containment management, cloning, fine-grained notifications management, object graph comparison and diff/merge, visiting patterns, clipboard management, validation, support for \emph{design by contract} by integrating assertions from Java Modeling Language (JML), Metaprogramming, Aspect Oriented Programming. \ab{Experimented programmers may extend PAMELA defining new annotations.}

%A strong interest of the approach is that the model is encoded in Java, and must be compiled. It forces the Java compiler to perform required checks %for a PAMELA model encoded in a strong typed program. Execution semantics of model is fully compatible with Java semantics. Many validation rules %are automatically performed through classical Java compilation, independently of underlying PAMELA execution semantics.

@@

Listing \ref{lst:model} shows a very basic model with two entities: \emph{Book} and \emph{Library}. Entity \emph{Book} defines two read-write single properties \emph{title} and \emph{ISBN} with single cardinality and with \texttt{String} type. Entity \emph{Book} also defines a constructor with initial \emph{title} value. Entity \emph{Library} defines a read-write multiple properties \emph{books} referencing \emph{Book} instances. Note that this code is sufficient to execute the model, while no additional line of code is required (only Java interfaces and API methods are declared here).

@@

The lines 3--4 instantiates a \texttt{ModelContext} by introspecting and computing the closure of concepts graph obtained while starting from \texttt{Library} entity and following \texttt{parentEntities} and \texttt{properties} relationships. This call builds at runtime a \emph{PAMELAModel}, while dynamically following links reflected by compiled byte-code. A factory \texttt{ModelFactory} is then instantiated using that \texttt{ModelContext}, allowing to create \emph{Library} and \emph{Book} instances.

Custom code can be easily added to this model as we show in Listing \ref{lst:custom}. It shows how to integrate custom code to the fully interpreted \emph{Book} entity described above. The partial custom implementation is offered by a partial class (note the \texttt{abstract} keyword), declared in the annotation header of model entity. Custom implementations are defined using classical Java implementation/overrides scheme. Here we define the implementation of the \texttt{read()} method, which has no annotation (and thus, can not be processed by the PAMELA framework), and also the implementation of a custom getter for \emph{title}, returning a default value when no value is defined for that property. Note that this implementation references a default interpretated implementation (call to \texttt{performSuperGetter(String)} method).

### 2021-02-02 implementation.tex
work on architecture and process (AB)

The PAMELA implementation is available online. \footnote {https://github.com/openflexo-team/pamela}

@@

The package \texttt{org.openflexo.pamela} contains the following feature-related java interfaces:
\begin{itemize}
    \item \texttt{AccessibleProxyObject} is the interface that PAMELA objects should extend in order to benefit from base features such as generic default implementation, containment management, notification, object graph comparison and diff/merge, visiting patterns, etc.
    \item \texttt{CloneableProxyObject} exposes features related to cloning.
    \item \texttt{DeletableProxyObject} exposes features related to deletion management.
    \item \texttt{SpecifiableProxyObject} exposes dynamic assertion checking features in the context of JML (contract management) use. 
\end{itemize}

@@

From a technical point of view, PAMELA implementation uses the \emph{javassist} reflection library (see \cite{shigueru2000}) which provides the \texttt{MethodHandler} mechanism, which is a way to override the java dynamic binding. Invoking a method on an object which is part of a PAMELA model, causes the real implementation to be called when existing (more precisely dispatch code execution between all provided implementations), or the required interpretation according to underlying model to be executed. This provides also an extension point allowing to instrument the code, which is used for other features such as undo/redo stack management, and assertion checking at run-time (support for Design by Contract, aka JML).

The PAMELA framework is a 100\% pure java (\textgreater 1.5), compilable by a classical java compiler and executable in a classical Java Virtual Machine.

>>>>

The PAMELA implementation is available online\footnote {\url{https://github.com/openflexo-team/pamela}}.

@@

The package \texttt{org.openflexo.pamela} contains the following feature-related Java interfaces:
\begin{itemize}
    \item \texttt{AccessibleProxyObject} is the interface that PAMELA objects should extend in order to benefit from base features such as generic default implementation, containment management, notification, object graph comparison and diff/merge, visiting patterns, etc.
    \item \texttt{CloneableProxyObject} exposes features related to cloning.
    \item \texttt{DeletableProxyObject} exposes features related to deletion management.
    \item \texttt{SpecifiableProxyObject} exposes dynamic assertion checking features in the context of JML (contract management) use. 
\end{itemize}

@@

From a technical point of view, PAMELA implementation uses the \emph{javassist} reflection library (see \cite{shigueru2000}) which provides the \texttt{MethodHandler} mechanism, which is a way to override the Java dynamic binding. Invoking a method on an object which is part of a PAMELA model, causes the real implementation to be called when existing (more precisely dispatch code execution between all provided implementations), or the required interpretation according to underlying model to be executed. This provides also an extension point allowing to instrument the code, which is used for other features such as undo/redo stack management, and assertion checking at run-time (support for Design by Contract, aka JML).

The PAMELA framework is a 100\% pure Java (\textgreater 1.5), compilable by a classical Java compiler and executable in a classical Java Virtual Machine.

\subsection{PAMELA code base figures}

\todo{Surely you can say more about its structure, size of components, ... other interesting and relevant observations on the codebase?}

### 2021-02-02 pamelaProcess.tex
work on architecture and process (AB)

We advocate for a strong coupling between model and source-code, to give architects and developers a way to both interact during the whole development cycle. PAMELA is an annotation-based Java modeling framework providing a smooth integration between model and code, without code generation nor externalized model serialization. The idea is to avoid separation between modeling and code to facilitate consistency management and to /avoid round-trip issues.

\subsection{PAMELA Modeling/Development Process}

Coupling model and code into the same artifact opens new ways of programming. The classical (metadata enabled) programming process relies on \emph{programmers} that produce code reusing pre-existing modeling concepts. These concepts are implemented by \emph{modelers} that provide the right annotations the programmers use. This is, for instance, the process followed by Jakarta EE (JEE) developers reusing JEE specific annotations. The evolution rhythm between models and code is low. This programming way is still possible with PAMELA, but we allow the ability to reach a high evolution rhythm when the programmer becomes also the modeler. In fact, when a pattern, an abstraction, a generalization is identified by the programmer, s/he can use PAMELA to develop and capitalize on this abstraction by increasing PAMELA metamodel.

@@

The developed metamodels are implemented by annotations that rely on Java/JVM entities and mechanisms. They include consistency checkings, which constrain their use and helps the programmer to avoid inconsistencies or errors. We have experimented their use with setter/getter to define POJO entities, with traits to implement multiple inheritance or with roles and rules to set security rules on classes.

@@

>>>>

We advocate for a strong coupling between model and source-code, to give architects and developers a way to both interact during the whole development cycle. PAMELA is an annotation-based Java modeling framework providing a smooth integration between model and code, without code generation nor externalized model serialization. The idea is to avoid separation between modeling and code to facilitate consistency management and to /avoid round-trip issues. Figure \ref{fig:PamelaVision} summarizes the PAMELA architecture.

\begin{figure}
    \centering
    \includegraphics[width=1.0 \columnwidth]{PamelaVisionV2.pdf}
    \caption{PAMELA approach for modelling}
    \label{fig:PamelaVision}
\end{figure}

%\subsection{PAMELA Modeling/Development Process}

Coupling model and code into the same artifact opens new ways of programming. The classical (metadata enabled) programming process relies on \emph{programmers} that produce code reusing pre-existing modeling concepts. These concepts are implemented by \emph{modelers} that provide the right annotations the programmers use. This is, for instance, the process followed by Jakarta EE (JEE) developers reusing JEE specific annotations. The evolution rhythm between models and code is low. This programming way is still possible with PAMELA, but we allow the ability to reach a high evolution rhythm when the programmer also becomes the modeler. In fact, when a pattern, an abstraction, a generalization is identified by the programmer, s/he can use PAMELA to develop and capitalize on this abstraction by increasing PAMELA metamodel.

@@

The developed metamodels are implemented by annotations that rely on Java/JVM entities and mechanisms. They include consistency checkings, which constrain their use and helps the programmer to avoid inconsistencies or errors. We have first experimented their use with setter/getter to define POJO entities, with traits to implement multiple inheritance or with roles and rules to set security rules on classes.

@@

\ab{In this presentation, we focus on the programming use of PAMELA, when programmers reuse existing annotations. The full power of PAMELA arises when programmers become modelers defining their own abstractions/annotations.}
